---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r echo=FALSE}
library(knitr)
opts_chunk$set(echo=TRUE, fig.path='./figure/')
setwd("C:/Users/Axel Schwanke/Documents/GitHub/RepData_PeerAssessment1")
```

## Introduction

This assignment makes use of data from a personal activity monitoring device. 
This device collects data at 5 minute intervals through out the day
(eg. interval 1455 means 14:55 o'clock until 15:00 o'clock (24h time)).
The data consists of two months of data from an anonymous individual collected 
during the months of October and November, 2012 and include the number of steps 
taken in 5 minute intervals each day.

Dataset:
https://d396qusza40orc.cloudfront.net/repdata/data/activity.zip


## Loading and preprocessing the data
```{r}
if (!file.exists("./activity.csv")) {
    unzip(zipfile="./activity.zip") # needs only to be unzipped once
}
data <- read.csv(file="./activity.csv")
str(data)
```
Columns 'steps' and 'interval' have format 'int', which we can work on.
The format of the column 'date' is 'factor' and should be 'date'.

Transforming format of column 'date' from 'factor' to 'date'
```{r}
data$date <- as.Date(data$date)
summary(data)
```



## What is mean total number of steps taken per day?

#### Total number of steps taken per day
```{r}
total_steps_per_day <- tapply(data$steps, data$date, sum)
summary(total_steps_per_day)
```
The total number of steps per day varies between 41 and 21190.
At 8 days no steps were recorded (NA values).

#### Histogram of the total number of steps taken each day
```{r fig.height=4.0, fig.width=8.0}
hist(total_steps_per_day, breaks=20, xlab="Total number of steps per day", 
     ylab="Percentage", main="Total number of steps taken per day")
```


#### The mean and median number of steps taken per day
```{r}
mean_steps_per_day <- mean(total_steps_per_day, na.rm=TRUE)
median_steps_per_day <- median(total_steps_per_day, na.rm=TRUE)
```
The mean is `r format(mean_steps_per_day, scientific=FALSE)` 
and the median is `r format(median_steps_per_day, scientific=FALSE)`.

The mean and median have about the same value. 
The distribution is fairly symmetric and not skewed.



## What is the average daily activity pattern?

#### Average of number of steps in each 5 minute interval over all days
```{r}
avg_steps_per_interval <- tapply(data$steps, data$interval, mean, na.rm=TRUE)
intervals <- sort(unique(data$interval))
```

#### Time-series plot 

The following figure shows the time-series plot of the 5-minute interval (x-axis) 
and the average number of steps taken, averaged across all days (y-axis).
```{r}
plot(intervals, avg_steps_per_interval, type="l", lwd=2,
     xlab="Interval", ylab="Mean number of steps", main="Average daily activity")
```

#### Most active interval

```{r}
mostActiveInterval <- intervals[which.max(avg_steps_per_interval)]
```
The interval `r mostActiveInterval` contains the maximum number of steps on average.
It is the interval from 8:35 to 8:40 o'clock in the morning.



## Imputing missing values

```{r}
sumNA <- sum(is.na(data$steps))
```
The total number of missing values is `r sumNA`.

A common approach is to fill missing values with the average of non-missing values of that interval.

```{r}
# create a new dataset - initalize
dataNew <- data
# loop over all rows
for (i in 1:nrow(data)) {
    # if column steps contains NA
    if (is.na(data$steps[i])) {
        # replace NA with the mean of all non-NA values of the same interval
        dataNew$steps[i] <- mean(data$steps[data$interval == data$interval[i]],
                                 na.rm=TRUE)
    }
}
```

#### Histogram of the total number of steps taken each day

```{r fig.height=4.0, fig.width=8.0}
total_steps_per_dayNew <- tapply(dataNew$steps, dataNew$date, sum)

hist(total_steps_per_dayNew, breaks=20, xlab="Total number of steps per day", 
     ylab="Percentage", main="Total number of steps taken per day")
```
```

#### The mean and median number of steps taken per day
```{r}
mean_steps_per_dayNew <- mean(total_steps_per_dayNew)
median_steps_per_dayNew <- median(total_steps_per_dayNew)
```
The mean is `r format(mean_steps_per_dayNew, scientific=FALSE)` 
and the median is `r format(median_steps_per_dayNew, scientific=FALSE)`.

Mean and median are the same. 
The mean is the same as without the imputed values, but the median has changed.
When imputing missing values ... ???


## Are there differences in activity patterns between weekdays and weekends?

We first need to calculate whether a specific data is on a weekday or on a weekend.
```{r fig.height=9.0, fig.width=8.0}
# set local time - otherwise getting German Samstag, Sonntag
Sys.setlocale("LC_TIME", "English")

# create a new factor variable weekend
wd <- weekdays(as.Date(data$date))
dataNew$weekend <- FALSE
dataNew$weekend <- (wd == "Saturday" | wd == "Sunday")
dataNew$weekend <- as.factor(dataNew$weekend)

weekend_mean <- tapply(dataNew$steps[dataNew$weekend==TRUE], 
                       dataNew$interval[dataNew$weekend==TRUE], mean)
weekday_mean <- tapply(dataNew$steps[dataNew$weekend==FALSE], 
                       dataNew$interval[dataNew$weekend==FALSE], mean)

max_steps <- ceiling(max(weekday_mean,weekend_mean))
par(mfrow=c(2,1))
plot(intervals, weekday_mean, type="l", lwd=2, ylim=c(0,max_steps),
     xlab="Interval", ylab="Avg. number of steps", main="weekdays")
plot(intervals, weekend_mean, type="l", lwd=2, ylim=c(0,max_steps),
     xlab="Interval", ylab="Avg. number of steps", main="weekend days")
par(mfrow=c(1,1))
```

As one can clearly recognize, that on weekdays the most active time is in the 
morning between about 5:30 o'clock and 10:00 o'clock.
At weekends the active time starts later, but is more spread out between 
8:00 o'clock in the morning and 20:00 o'clock in the evening.

```{r}
weekday_sum <- sum(weekday_mean)
weekend_sum <- sum(weekend_mean)
```

One can also see that the mean sum of steps taken at weekends with 
`r format(round(weekend_sum), scientific=FALSE)` is substantially higher than 
at weekdays with `r format(round(weekday_sum), scientific=FALSE)` .


